Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex

\begin{table}
\centering
\begin{tabular}{l|l|l} \hline \textbf{Property} &
\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
no descendant besides the sequence causing insertion of the node) & 2,601
(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
(including sample metadata, imputed Pango lineage assignments, and results of
HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
\end{tabular}
\end{table}

@@

\subsection{Computational resources}

@@

For validation purposes, we compare the genetic relationships in the
sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
restricted to non-recombinant lineages. To allow for comparison, the ARGs were
"simplified" \citep{Kelleher2018-xc} to retain only those tips of our genealogy
that match the samples present in a specific NextStrain tree. These simplified
"backbone" ARGs exhibit very little recombination (6 and 5 trees across the
genome for the wide and long ARGs, respectively).

@@

\subsection{Putative recombination events}

@@

\subsection{Early recombinants reported by Jackson et al. (2021)}

\subsection{Pango recombinants}

The wide ARG only contains samples from three Pango-designated recombinant
lineages (XA, XB, and XC), due to the fact that it only includes samples up to
mid-2021. By contrast, the long ARG (subsampled to mid-2022) contains samples
from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table
3). For Pango recombinant analysis, we therefore focus on the long ARG.

In the wide ARG, every sequence of the three Pango recombinants trace back to
some recombination node. XA, XB, and XC.

In the unfiltered long ARG, the sequences of 28 Pango recombinants trace back
to some recombinant node(s); however, the sequences of the five Pango
recombinants XP, XU, XAA, XAG, and XAK do not. A likely explanation for the XP
recombinant sequences not having a recombinant origin in the long ARG is that
its characteristic multibase deletion
(https://github.com/cov-lineages/pango-designation/issues/481) was masked
during preprocessing and therefore was not incorporated into the HMM. Upon
tracing the origin of XAG through the ARG, XAG might have been caused by
incorrect imputation of its parent nodes(?). XAK likely does not appear to have
a recombinant origin because of random subsampling, but it may with a more
complete tree.

The sequences of 14 Pango recombinants trace back to a single recombination
node. The sequences of the remaining 18 Pango recombinants trace back to
multiple recombination nodes, however. This may be a result of undersampling in
the long ARG, which was built by randomly subsampling at most 1,000 sequences
from each daily batch. Having extra recombination nodes in the long ARG
appeared to be a more parsimonious explanation for the 19 Pango recombinants
than having a single recombination node with many immediately ancestral
mutations (including recurrent mutations and reversions), despite that the
recombinants may indeed trace to a single origin.

The mosaic genome structure of all the 32 Pango recombinants is concordant with
the structure determined by the Pango Lineage Designation Committee
(Supplementary Table 3). For example, the genome of XD (also known as
``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

@@

Next, we examine recombination nodes in the filtered long ARG (sampled to
mid-2022) that represent previously unnamed putative recombinant sequences,
which do not have a Pango recombinant designation. For this proof-of-concept
study, we have arbitrarily picked 12 recombination nodes that seem plausible
(Supplementary Table 4). Coincidentally, all these nodes involve Omicron
subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
inserted into the long ARG on or after January 8, 2022; (2) have at least 10
descendent samples; and (3) have no mutational differences (including immediate
reversions) from their parent nodes.

@@

\subsection{Recombinant origins of the Pango lineages XA and XAG}

XX

We explore recombination signals across the SARS-CoV-2 genome by inspecting the
breakpoint intervals of the filtered recombination nodes from the wide ARG
(sampled to mid-2021) (Figure 4). The length of a breakpoint interval reflects
uncertainty in the estimated location of a breakpoint. On average, the length
of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
23,666 bases). This broad distribution of interval length indicates that the
location of a breakpoint often could not be precisely determined given the
observed genetic variation.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra}
Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
Mutations are categorised by type (i.e., ancestral state >
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C > U) and the
lighter colours the reverse (e.g., U > C).}
\end{figure}

\subsection{Mutations under selection}

XX

>>>>

\begin{table}
\centering
\begin{tabular}{l|l|l} \hline \textbf{Property} &
\textbf{Wide ARG} & \textbf{Long ARG} \\ \hline Last sampling date & 2021-06-30
& 2022-06-30 \\ Local trees & 1,496 & 958 \\ \hline
\multicolumn{3}{l}{\textit{Nodes}} \\ All nodes (sample and non-sample) &
1,453,347 & 783,231 \\ Sample nodes & 1,265,685 (87\%) & 657,239 (84\%) \\ Tree
adjustment nodes (excluding recombination nodes) & 80,830 (6.34\%) & 72,107
(10.97\%) \\ Recombination nodes & 4,123 (0.28\%) & 2,078 (0.27\%) \\ \hline
\multicolumn{3}{l}{\textit{Types of recombination nodes}} \\ Singletons (i.e.
no descendant besides the sequence causing insertion of the node) & 2,601
(63\%) & 1,315 (63\%) \\ Two-parent recombinants & 3,521 (85\%) & 1,869 (90\%)
\\ HMM-consistent recombinants & XX (XX\%) & XX (XX\%) \\ \hline
\multicolumn{3}{l}{\textit{Mutations}} \\ Total mutations & 1,213,193 &
1,062,072 \\ Mutations on a terminal edge & 758,903 (62.55\%) & 767,111
(72.23\%) \\ Recurrent mutations & 74,719 (6.16\%) & 50,099 (4.72\%) \\
Reversions & 72,617 (5.99\%) & 48,226 (4.54\%) \\ Immediate reversions (one
intervening node only) & 3,541 (0.29\%) & 1,347 (0.13\%) \\ \hline
\multicolumn{3}{l}{\textit{Storage and loading time}} \\ tszipped file
(including sample metadata, imputed Pango lineage assignments, and results of
HMM runs for recombination nodes) & ~57 Mb & ~36 Mb \\ Bytes to encode one
sample & 8.29 & 10.83 \\ Load time (seconds) & 0.9 & 0.5 \\ \hline
\end{tabular}
\caption{\label{tab:args}Summary of the inferred ARGs.
}
\end{table}

@@

@@

For validation purposes, we compare the genetic relationships in the
sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
restricted to non-recombinant lineages. To allow for comparison, the ARGs were
``simplified'' \citep{Kelleher2018-xc} to retain only those samples of our genealogy
that match those present in a specific NextStrain tree. These simplified
``backbone'' ARGs exhibit very little recombination (6 and 5 trees across the
genome for the wide and long ARGs, respectively).

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra}
Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to \cite{Yi2021-sc}.
Mutations are categorised by type (i.e., ancestral state $>$
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C$>$U) and the
lighter colours the reverse (e.g., U$>$C).}
\end{figure}

% If we do this, goes here
% \subsection{Mutations under selection}
% XX

\subsection{Early recombinants reported by Jackson et al. (2021)}
% - Early Recombinants from Jackson et al
%     Discuss results from Wide ARG in detail. Go into the details of the
% forward-backward HMM runs for each example. Be much more explicity about the
% single origin for the different groups (maybe give one of Ana's graphs for
% Group A (XA I think?)

% - Distribution of breakpoint intervals
%     Study the breakpoint intervals in the Wide ARG. Because we are interested
% in the intervals, we want to be sure that these are robustly reported by the
% HMM. Hence, we focus on those that have two parents in the mirrored and
% unmirrored HMM runs, the same set of mutations, and the left and right parents
% are consistently imputed for Pango lineage status.
%     We focus on the Wide ARG here because it covers (roughly?) the same time
% period as the RIPPLES analysis.

We explore recombination signals across the SARS-CoV-2 genome by inspecting the
breakpoint intervals of the filtered recombination nodes from the wide ARG
(sampled to mid-2021) in Figure~\ref{fig:breakpoint-distribution}.
The length of a breakpoint interval reflects
uncertainty in the estimated location of a breakpoint. On average, the length
of the breakpoint intervals is 2,895 bases (median, 1,670 bases; range, 1 to
23,666 bases). This broad distribution of interval length indicates that the
location of a breakpoint often could not be precisely determined given the
observed genetic variation.

\begin{figure}
\centering
% \includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:breakpoint-distribution}
Distribution of recombination breakpoints and mutations along the genome.}
\end{figure}

\subsection{Divergence between recombinant parents}
% - Divergence between recombinant parents
%     We study the recombination events in the Long ARG in terms. A recombinant
% will not always have two parents, for reasons X, Y and Z. As a first
% approximation, we can assume that each breakpoint corresponds to a
% recombination event (although there are some exceptions). Here we have to get
% into all the hairy details about which sets of breakpoints we filter and don't
% filter.

@@

\subsection{Recombinant Pango lineages}

In this section we examine the Long ARG to determine how well it reflects
the structure of previously identified recombinant lineages.
There are [high level summary of the pango recombinant lineages and samples
in the Long ARG].
long ARG (subsampled to mid-2022) contains samples
from 33 Pango-designated recombinant lineages (XA to XAK; Supplementary Table)
[Probably could have a table here to summarise?]

The sequences of 28 Pango recombinants trace back
to some recombinant node(s); however, the sequences of the five Pango
% jk: really? I thought it was just XP that wasn't inferred to be a
% recombinant.
recombinants XP, XU, XAA, XAG, and XAK do not.
A likely explanation for the XP
recombinant sequences not having a recombinant origin in the long ARG is that
its characteristic multibase deletion
(\url{https://github.com/cov-lineages/pango-designation/issues/481}) was masked
during preprocessing and therefore was not incorporated into the HMM.
% Eh?
Upon tracing the origin of XAG through the ARG, XAG might have been caused by
incorrect imputation of its parent nodes(?). XAK likely does not appear to have
a recombinant origin because of random subsampling, but it may with a more
complete tree.

[FIXME this isn't quite right - not all the sequences for a given pango
lineage trace back to the same recombination event, so won't have the same
intervals.]
The mosaic genome structure of all the 32 Pango recombinant lineages is concordant with
the structure determined by the Pango Lineage Designation Committee
(Supplementary Table 3). For example, the genome of XD (also known as
``Deltacron'') is inferred to consist of B.1.617.2 (Delta) on the 5’ end, BA.1.17
(Omicron) in the middle, and B.1.617.2 (Delta) on the 3’ end.

The wide ARG contains samples from three Pango-designated recombinant
lineages (XA, XB, and XC) and N samples, since it only includes data
up to mid-2021.
In the wide ARG, every sequence of the three Pango recombinants trace back to
some recombination node. XA, XB, and XC.

For the long ARG, we examine the lineages in increasing complexity of the
inferred evolutionary relationships between samples assigned to Pango
X lineages.

% - Recombinant Pango lineages
%     How well do we recover recombinant Pango lineages? Work through the
% examples in increasing compexity, using headings to split things up if
% necessary. (i.e., Single origin, Multiple origin, Interellated lineages). We'd
% need to be clear about what these terms mean, of course. We could include
% lineage graphs for all of these examples?

% OLD TEXT

% The sequences of 14 Pango recombinants trace back to a single recombination
% node. The sequences of the remaining 18 Pango recombinants trace back to
% multiple recombination nodes, however. This may be a result of undersampling in
% the long ARG, which was built by randomly subsampling at most 1,000 sequences
% from each daily batch. Having extra recombination nodes in the long ARG
% appeared to be a more parsimonious explanation for the 19 Pango recombinants
% than having a single recombination node with many immediately ancestral
% mutations (including recurrent mutations and reversions), despite that the
% recombinants may indeed trace to a single origin.

\subsubsection{Single origin}

We would usually assume that all sequences assigned to a given Pango X
lineage are descendants of a single recombinant sequence, arising as a
result of a mixed infection followed by onward transmission. Thus, we
would also expect our ARG to reflect this evolutionary history, where
we have a all the samples assigned to this recombinant lineage tracing
back to a single recombination node, representing this original
recombination event.

The sequences of 14 Pango recombinants trace back to a single recombination
node. The details are given in Figure X, where we [give the details of
the numbers of samples, etc for each of these 14 lineages] and
show the lineage subgraph for X?, which makes this simple structure
clear. [Explain the lineage subgraph format a bit]

\subsubsection{Multiple origin}
The ARG inference has no pre-defined knowledge of Pango X lineage assignments,
and there is therefore no particular requirement that all the samples
assigned to a given lineage must trace back to a single recombination
event. In this section we examine the Pango X lineage samples that
are inferred to have multiple recombinant origins (i.e., trace
back to different recombination nodes in the ARG), but are still
form monophyletic clades % is this the right term?
descending from these nodes .

In Figure XXX we give the details of these multiple pango lineages,
along with an example lineage subgraph.

Interestingly, XC is inferred to be single origin in the Wide ARG but
not in the long ARG. See lineage subgraphs in supplemental figure Y.
This suggests that insufficient sampling may be the explanation for
many of these multiple origin events.

\subsubsection{Complex relationships between X lineages}

Some lineages are inferred to have overlapping ancestral histories.
Discuss XAG example in Figure XXX.

@@

Next, we examine recombination nodes in the long ARG (sampled to
mid-2022) that represent previously unnamed putative recombinant sequences,
which do not have a Pango recombinant designation. For this proof-of-concept
study, we have arbitrarily picked 12 recombination nodes that seem plausible
(Supplementary Table 4). All these nodes involve Omicron
subvariants (BA.1, BA.2, BA.4, and BA.5). These recombination nodes (1) were
inserted into the long ARG on or after January 8, 2022; (2) have at least 10
descendent samples; and (3) have no mutational differences (including immediate
reversions) from their parent nodes.

@@
